Hypercoercions and a Framework for Equivalence of Cast Calculi

### 2019-10-24 paper.tex
revise sec5

\newcommand{\kOOmt}[0]{\mathtt{stop}}
\newcommand{\kOOconsI}[5]{
	[\mathtt{cons}^{\POOprod{#1}{#2}} \; \square \; \langle#3,#4\rangle ]#5}
\newcommand{\kOOconsII}[4]{
	[\mathtt{cons}^{\POOprod{#1}{#2}} \; #3 \; \square]#4}
\newcommand{\kOOinl}[3]{[\mathtt{inl}^{\POOsum{#1}{#2}}\; \square]#3}
\newcommand{\kOOinr}[3]{[\mathtt{inr}^{\POOsum{#1}{#2}}\; \square]#3}
\newcommand{\kOOappI}[3]{
  [\square \; \langle#1,#2\rangle]#3
}
\newcommand{\kOOappII}[2]{
  [#1 \; \square]#2}
\newcommand{\kOOcar}[1]{[\mathtt{fst} \; \square]#1}
\newcommand{\kOOcdr}[1]{[\mathtt{snd} \; \square]#1}
\newcommand{\kOOcaseI}[4]{
  [\mathtt{case} \; \square \; \langle#1,#3\rangle \; \langle#2,#3\rangle ]#4}
\newcommand{\kOOcaseII}[4]{
  [\mathtt{case} \; #1 \; \square \; \langle#2,#3\rangle ]#4}
\newcommand{\kOOcaseIII}[3]{
  [\mathtt{case} \; #1 \; #2 \; \square]#3}
\newcommand{\kOOcast}[2]{
  [\square \langle #1 \rangle]#2}

%\newcommand{\kOOmt}[0]{\mathtt{stop}}
%\newcommand{\kOOconsI}[3]{\mathtt{cons_1} \; #1 \; #2 \; #3}
%\newcommand{\kOOconsII}[2]{\mathtt{cons_2} \; #1 \; #2}
%\newcommand{\kOOinl}[1]{\mathtt{inl} \; #1}
%\newcommand{\kOOinr}[1]{\mathtt{inr} \; #1}
%\newcommand{\kOOappI}[3]{
% \mathtt{app_1} \; #1 \; #2 \; #3
%}
%\newcommand{\kOOappII}[2]{
% \mathtt{app_2} \; #1 \; #2}
%\newcommand{\kOOcar}[1]{
% \mathtt{fst} \; #1}
%\newcommand{\kOOcdr}[1]{
% \mathtt{snd} \; #1}
%\newcommand{\kOOcaseI}[4]{
% \mathtt{case_1} \; #1 \; #2 \; #3 \; #4}
%\newcommand{\kOOcaseII}[4]{
% \mathtt{case_2} \; #1 \; #2 \; #3 \; #4}
%\newcommand{\kOOcaseIII}[3]{
% \mathtt{case_3} \; #1 \; #2 \; #3}
%\newcommand{\kOOcast}[2]{
% \langle #1 \rangle #2}

\newcommand{\typingHC}[3]{#1 : #2 \Longrightarrow #3}
\newcommand{\hcvOOinj}[2]{\mathtt{Dyn}_{#1}(#2)}
%\newcommand{\hcvOOfun}[5]{\mathtt{fun} \; #1 \; #2 \; #3 \; #4 \; #5}
\newcommand{\hcvOOfun}[5]{\mathtt{fun} \; #2 \; #1 \; #3 \; #4 \; #5}
\newcommand{\hcvOOtt}[0]{\mathtt{unit}}
%\newcommand{\hcvOOcons}[4]{\mathtt{cons}\;#1\;#2\;#3\;#4}
\newcommand{\hcvOOcons}[4]{\mathtt{cons}\;#1\langle#2\rangle\;#3\langle#4\rangle}
%\newcommand{\hcvOOinl}[2]{\mathtt{inl}\;#1\;#2}
%\newcommand{\hcvOOinr}[2]{\mathtt{inr}\;#1\;#2}
\newcommand{\hcvOOinl}[2]{\mathtt{inl}\;#1\langle#2\rangle}
\newcommand{\hcvOOinr}[2]{\mathtt{inr}\;#1\langle#2\rangle}
\newcommand{\hckOOmt}[0]{\mathtt{stop}}
\newcommand{\hckOOconsI}[3]{\mathtt{cons_1}\;#1\;#2\;#3}
\newcommand{\hckOOappII}[2]{\mathtt{app_2}\;#1\;#2}
\newcommand{\sidecond}[1]{\text{if}\;#1}
% Lazy D cast calculus on space-inefficient CEK
\newcommand{\judgeCreduce}[2]{#1 \longmapsto_{\mathcal{x}} #2}
\newcommand{\judgeCreduceTrans}[2]{#1 \longmapsto_{\mathcal{x}}^{*} #2}
\newcommand{\judgeCeval}[2]{eval_{\mathcal{x}}(#1) = #2}
\newcommand{\redrule}[3]{#1 & \longmapsto_\mathcal{x} & #2 & #3\\}
\newcommand{\hiredrule}[3]{\highlight{#1} & \highlight{\longmapsto_\mathcal{x}} 
& \highlight{#2} & \highlight{#3} \\}
% blame calculus on space-efficient CEK
\newcommand{\judgeSreduce}[3]{#2 \longmapsto_{\mathcal{S}(#1)} #3}
\newcommand{\judgeSreduceTrans}[3]{#2 \longmapsto_{\mathcal{S}(#1)}^{*} #3}
\newcommand{\judgeSeval}[3]{eval_{\mathcal{S}(#1)}(#2) = #3}
\newcommand{\redruleS}[3]{#1 & \longmapsto_{\mathcal{S}(C)} & #2 & #3\\}
\newcommand{\hiredruleS}[3]{\highlight{#1} & 
\highlight{\longmapsto_{\mathcal{S}(C)}} & \highlight{#2} & \highlight{#3} \\}
% Normal Coercion
\newcommand{\ncProj}[2]{#1?^{#2}}
\newcommand{\ncInj}[1]{#1!}
\newcommand{\ncId}[0]{\iota}
\newcommand{\ncSeq}[2]{#1;#2}
\newcommand{\ncFail}[1]{\bot^{#1}}
\newcommand{\ncFun}[2]{\POOfun{#1}{#2}}
\newcommand{\ncProd}[2]{\POOprod{#1}{#2}}
\newcommand{\ncSum}[2]{\POOsum{#1}{#2}}
% Hypercoercion
\newcommand{\hyperCoercionI}[0]{\mathtt{id\star}}
\newcommand{\hyperCoercionC}[3]{#1 \overset{#2}{\curvearrowright} #3}
% machine state simulations
\newcommand{\eqvS}[4]{#3 \approx_{\mathcal{S}\mathcal{S}} #4}
\newcommand{\eqvSD}[3]{#2 \approx_{\mathcal{SD}} #3}
% to-dos
\newcommand{\todoKC}[1]{\todo[inline]{KC needs to #1}}
\newcommand{\todoKCFixed}[0]{\todo[inline]{Fixed. -KC}}
% abbrev
\newcommand{\castCalculus}[0]{$\lambda_{\rightarrow}^{\langle\cdot\rangle}$}
% names
\newcommand{\ineffCEK}{$\mathcal{C}$}
\newcommand{\ineffCEKD}{$\mathcal{D}$}
\newcommand{\ineffCEKUD}{$\mathcal{UD}$}
\newcommand{\judgeDreduce}[2]{#1 \longmapsto_{\mathcal{D}} #2}
\newcommand{\judgeDreduceTrans}[2]{#1 \longmapsto_{\mathcal{D}}^{*} #2}
\newcommand{\judgeDeval}[2]{eval_{\mathcal{D}}(#1) = #2}
\newcommand{\judgeUDreduce}[2]{#1 \longmapsto_{\mathcal{UD}} #2}
\newcommand{\judgeUDreduceTrans}[2]{#1 \longmapsto_{\mathcal{UD}}^{*} #2}
\newcommand{\judgeUDeval}[2]{eval_{\mathcal{UD}}(#1) = #2}
\newcommand{\effCEK}[1]{$\mathcal{S}(#1)$}
\newcommand{\evalEqv}[2]{\ensuremath{eval_{\text{#1}} = eval_{\text{#2}}}}
\newcommand{\continue}[2]{cont(#2,#1)}
\newcommand{\highlight}[1]{{\color{red} #1}}

@@

@@

\section{Definition of Hypercoercions} \label{sec:hypercoercion-definition}
\todo[inline]{
The figures of applyCast for hypercoercions is before coercions, even before 
the syntax definition of hypercoercions! -KC
}

@@

\begin{proposition}[\lazyD{} Hypercoercions Form a Monoid]
  \label{thm:hc-monoid}
  For all $c : T_1 \Longrightarrow T_2$,
  $c_1 : T_1 \Longrightarrow T_2$,
  $c_2 : T_2 \Longrightarrow T_3$, and
  $c_3 : T_3 \Longrightarrow T_4$,
  \begin{enumerate}
    \item $seq(id(T_1),c) = c$,
    \item $seq(c,id(T_2)) = c$, and
    \item $seq(seq(c_1, c_2), c_3) = seq(c_1, seq(c_2, c_3))$.
  \end{enumerate}
\end{proposition}
\begin{proof}
  Part (1) and (2) are by induction on $c$. Part (3) is by induction
  on $c_2$ and case analysis on $c_1$ then $c_3$.
\end{proof}

\begin{proposition}[\lazyD{} Identity Casts]
  For all $T$ and $l$, $cast(T,l,T) = id(T) $
\end{proposition}
\begin{proof}
  By induction on $T$.
\end{proof}

@@

Before turning to the definition of the abstract data types, we first
give the definition of values and cast results used by the \effCEK{C}, as they 
are mentioned in the definition of the ADTs. The $\hcvOOinj{P}{v}$ stands for 
values that are cast to the dynamic type, as in \citet{wadler2009well}.

@@

The \emph{Cast Abstract Data Type}, defined below, captures the set of
operators that a cast representation $C$ must provide for it to be
used with the \effCEK{C} machine.  The first three operators enables
\effCEK{C} to construct casts so we call them \textit{cast
  constructors}. The forth and last operator enables \effCEK{C} to
apply casts to values.

\begin{definition}[Cast Abstract Data Type (Cast ADT)]
  \label{def:cast-rep}
  A cast abstract data type is a set $Cast$, which is indexed by two types, 
  with four operators:
  \begin{description}
  \item[$id(T)$] constructs an identity cast from a type
  \item[$seq(c_1,c_2)$] composes two casts
  \item[$cast(T_1,l,T_2)$] constructs a cast from $T_1$ to $T_2$
  \item[$applyCast(v,c)$] applies a cast to a value, producing a result
  \end{description}
  We denote by $c : T_1 \Longrightarrow T_2$ to mean $c$ is in $Cast \; T_1 \; T_2$
  and we say ``$c$ is from $T_1$ to $T_2$''.
\end{definition}

@@

\figref{hc-applyCast} defines $applyCast(v,c)$ for \lazyD\ hypercoercion. 
Applying the identity cast for the 
dynamic type succeeds immediately. When applying a compound cast, we firstly 
apply the middle, then apply the tail. We denote by $r \; >>= \; f$ to mean 
that if $r$ is $\rOOsucc{v}$, the result is $f(v)$, otherwise the result 
is the failure.
$ applyMiddle(v,\ell,m)$ and $applyTail(v,t)$ are straightforward.
In the definition of $applyMiddle(v,\ell,m)$, we generalize 
shallow-consistency to compare middles and values ($v \smile m$) in the 
natural way.

@@

\begin{lemma}[\lazyD{} Hypercoercion is a \lazyD{} Cast ADT]
  \label{thm:hc-surely-lazyD}
\end{lemma}
\begin{proof} See the supplementary material. \end{proof}

\begin{theorem}[\lazyD{} Hypercoercion Respect \ineffCEKD]
  If $\judgetype{\emptyset}{e}{T}$ and $o : T$
  \[
  \text{\evalEqv{\effCEK{H}}{\ineffCEKD}}
  \]
\end{theorem}
\begin{proof}
  Immediately from Theorem~\ref{thm:surely-lazyD-correct} and 
  Lemma~\ref{thm:hc-surely-lazyD}.
  Alternatively, from Lemma~\ref{thm:surely-monoidic-eval},
  Lemma~\ref{thm:hc-surely-lazyD}, and 
  Proposition~\ref{thm:hc-monoid}.
\end{proof}

@@

>>>>

\newcommand{\kOOmt}[0]{\mathtt{stop}}
\newcommand{\kOOconsI}[5]{
	[\mathtt{cons}^{\POOprod{#1}{#2}} \; \square \; \langle#3,#4\rangle ]#5}
\newcommand{\kOOconsII}[4]{
	[\mathtt{cons}^{\POOprod{#1}{#2}} \; #3 \; \square]#4}
\newcommand{\kOOinl}[3]{[\mathtt{inl}^{\POOsum{#1}{#2}}\; \square]#3}
\newcommand{\kOOinr}[3]{[\mathtt{inr}^{\POOsum{#1}{#2}}\; \square]#3}
\newcommand{\kOOappI}[3]{
  [\square \; \langle#1,#2\rangle]#3
}
\newcommand{\kOOappII}[2]{
  [#1 \; \square]#2}
\newcommand{\kOOcar}[1]{[\mathtt{fst} \; \square]#1}
\newcommand{\kOOcdr}[1]{[\mathtt{snd} \; \square]#1}
\newcommand{\kOOcaseI}[4]{
  [\mathtt{case} \; \square \; \langle#1,#3\rangle \; \langle#2,#3\rangle ]#4}
\newcommand{\kOOcaseII}[4]{
  [\mathtt{case} \; #1 \; \square \; \langle#2,#3\rangle ]#4}
\newcommand{\kOOcaseIII}[3]{
  [\mathtt{case} \; #1 \; #2 \; \square]#3}
\newcommand{\kOOcast}[2]{
  [\square \langle #1 \rangle]#2}
\newcommand{\typingHC}[3]{#1 : #2 \Longrightarrow #3}
\newcommand{\hcvOOinj}[2]{\mathtt{Dyn}_{#1}(#2)}
%\newcommand{\hcvOOfun}[5]{\mathtt{fun} \; #1 \; #2 \; #3 \; #4 \; #5}
\newcommand{\hcvOOfun}[5]{\mathtt{fun} \; #2 \; #1 \; #3 \; #4 \; #5}
\newcommand{\hcvOOtt}[0]{\mathtt{unit}}
%\newcommand{\hcvOOcons}[4]{\mathtt{cons}\;#1\;#2\;#3\;#4}
\newcommand{\hcvOOcons}[4]{\mathtt{cons}\;#1\langle#2\rangle\;#3\langle#4\rangle}
%\newcommand{\hcvOOinl}[2]{\mathtt{inl}\;#1\;#2}
%\newcommand{\hcvOOinr}[2]{\mathtt{inr}\;#1\;#2}
\newcommand{\hcvOOinl}[2]{\mathtt{inl}\;#1\langle#2\rangle}
\newcommand{\hcvOOinr}[2]{\mathtt{inr}\;#1\langle#2\rangle}
\newcommand{\hckOOmt}[0]{\mathtt{stop}}
\newcommand{\hckOOconsI}[3]{\mathtt{cons_1}\;#1\;#2\;#3}
\newcommand{\hckOOappII}[2]{\mathtt{app_2}\;#1\;#2}
\newcommand{\sidecond}[1]{\text{if}\;#1}
% Lazy D cast calculus on space-inefficient CEK
\newcommand{\judgeCreduce}[2]{#1 \longmapsto_{\mathcal{x}} #2}
\newcommand{\judgeCreduceTrans}[2]{#1 \longmapsto_{\mathcal{x}}^{*} #2}
\newcommand{\judgeCeval}[2]{eval_{\mathcal{x}}(#1) = #2}
\newcommand{\redrule}[3]{#1 & \longmapsto_\mathcal{x} & #2 & #3\\}
\newcommand{\hiredrule}[3]{\highlight{#1} & \highlight{\longmapsto_\mathcal{x}} 
& \highlight{#2} & \highlight{#3} \\}
% blame calculus on space-efficient CEK
\newcommand{\judgeSreduce}[3]{#2 \longmapsto_{\mathcal{S}(#1)} #3}
\newcommand{\judgeSreduceTrans}[3]{#2 \longmapsto_{\mathcal{S}(#1)}^{*} #3}
\newcommand{\judgeSeval}[3]{eval_{\mathcal{S}(#1)}(#2) = #3}
\newcommand{\redruleS}[3]{#1 & \longmapsto_{\mathcal{S}(C)} & #2 & #3\\}
\newcommand{\hiredruleS}[3]{\highlight{#1} & 
\highlight{\longmapsto_{\mathcal{S}(C)}} & \highlight{#2} & \highlight{#3} \\}
% Normal Coercion
\newcommand{\ncProj}[2]{#1?^{#2}}
\newcommand{\ncInj}[1]{#1!}
\newcommand{\ncId}[0]{\iota}
\newcommand{\ncSeq}[2]{#1;#2}
\newcommand{\ncFail}[1]{\bot^{#1}}
\newcommand{\ncFun}[2]{\POOfun{#1}{#2}}
\newcommand{\ncProd}[2]{\POOprod{#1}{#2}}
\newcommand{\ncSum}[2]{\POOsum{#1}{#2}}
% Hypercoercion
\newcommand{\hyperCoercionI}[0]{\mathtt{id\star}}
\newcommand{\hyperCoercionC}[3]{#1 \overset{#2}{\curvearrowright} #3}
% machine state simulations
\newcommand{\eqvS}[4]{#3 \approx_{\mathcal{S}\mathcal{S}} #4}
\newcommand{\eqvSD}[3]{#2 \approx_{\mathcal{SD}} #3}
% to-dos
\newcommand{\todoKC}[1]{\todo[inline]{KC needs to #1}}
\newcommand{\todoKCFixed}[0]{\todo[inline]{Fixed. -KC}}
% abbrev
\newcommand{\castCalculus}[0]{$\lambda_{\rightarrow}^{\langle\cdot\rangle}$}
% names
\newcommand{\ineffCEK}{$\mathcal{C}$}
\newcommand{\ineffCEKD}{$\mathcal{D}$}
\newcommand{\ineffCEKUD}{$\mathcal{UD}$}
\newcommand{\judgeDreduce}[2]{#1 \longmapsto_{\mathcal{D}} #2}
\newcommand{\judgeDreduceTrans}[2]{#1 \longmapsto_{\mathcal{D}}^{*} #2}
\newcommand{\judgeDeval}[2]{eval_{\mathcal{D}}(#1) = #2}
\newcommand{\judgeUDreduce}[2]{#1 \longmapsto_{\mathcal{UD}} #2}
\newcommand{\judgeUDreduceTrans}[2]{#1 \longmapsto_{\mathcal{UD}}^{*} #2}
\newcommand{\judgeUDeval}[2]{eval_{\mathcal{UD}}(#1) = #2}
\newcommand{\effCEK}[1]{$\mathcal{S}(#1)$}
\newcommand{\evalEqv}[2]{\ensuremath{eval_{\text{#1}} = eval_{\text{#2}}}}
\newcommand{\continue}[2]{cont(#2,#1)}
\newcommand{\highlight}[1]{{\color{red} #1}}

@@

% URLs
\newcommand{\urlLazyDProof}{
	\url{https://github.com/LuKC1024/Blame-and-Coercion-Together-Again-for-the-Second-Time/tree/master/Proof}
}

@@

\section{Definition of Hypercoercions} \label{sec:hypercoercion-definition}

@@

\begin{proposition}[\lazyD{} Hypercoercions Form a Monoid]
  \label{thm:LazyD-HC-monoid}
  For all $c : T_1 \Longrightarrow T_2$,
  $c_1 : T_1 \Longrightarrow T_2$,
  $c_2 : T_2 \Longrightarrow T_3$, and
  $c_3 : T_3 \Longrightarrow T_4$,
  \begin{enumerate}
    \item $seq(id(T_1),c) = c$,
    \item $seq(c,id(T_2)) = c$, and
    \item $seq(seq(c_1, c_2), c_3) = seq(c_1, seq(c_2, c_3))$.
  \end{enumerate}
\end{proposition}
\begin{proof}
  Part (1) and (2) are by induction on $c$. Part (3) is by induction
  on $c_2$ and case analysis on $c_1$ then $c_3$.
\end{proof}

\begin{proposition}[\lazyD{} Identity Casts]
	\label{thm:LazyD-HC-identity-casts}
  For all $T$ and $l$, $cast(T,l,T) = id(T) $
\end{proposition}
\begin{proof}
  By induction on $T$.
\end{proof}

@@

Before turning to the definition of the abstract data types, we first
give the definition of values and cast results used by the \effCEK{C}, as they 
are mentioned in the definition of the ADTs. The $\hcvOOinj{I}{v}$ stands for 
values that are cast to the dynamic type, as in \citet{wadler2009well}.

@@

The \emph{Cast Abstract Data Type}, defined below, captures the set of
operators that a cast representation $C$ must provide for it to be
used with the \effCEK{C} machine.  The first three operators enable
\effCEK{C} to construct casts so we call them \textit{cast
  constructors}. The forth and last operator enables \effCEK{C} to
apply casts to values.

\begin{definition}[Cast Abstract Data Type (Cast ADT)]
  \label{def:cast-rep}
  A cast abstract data type is a set $Cast$, which is indexed by two types, 
  with four operators:
  \begin{description}
  \item[$id(T)$] constructs an identity cast from a type
  \item[$seq(c_1,c_2)$] composes two casts
  \item[$cast(T_1,l,T_2)$] constructs a cast from $T_1$ to $T_2$
  \item[$applyCast(v,c)$] applies a cast to a value, producing a cast result
  \end{description}
  We denote by $c : T_1 \Longrightarrow T_2$ to mean $c$ is in $Cast \; T_1 \; T_2$
  and we say ``$c$ is from $T_1$ to $T_2$''.
\end{definition}

@@

\figref{hc-applyCast} defines $applyCast(v,c)$ for \lazyD\ hypercoercions. 
Applying the identity cast for the 
dynamic type succeeds immediately. Otherwise, we firstly 
apply the middle, then apply the tail. We denote by $r \; >>= \; f$ to mean 
that if $r$ is $\rOOsucc{v}$, the result is $f(v)$, otherwise the result 
is the failure.
$ applyMiddle(v,\ell,m)$ and $applyTail(v,t)$ are straightforward.
In the definition of $applyMiddle(v,\ell,m)$, we generalize 
shallow-consistency to compare middles and values ($v \smile m$) in the 
natural way.

@@

\begin{lemma}[\lazyD{} Hypercoercion is a \lazyD{} Cast ADT]
  \label{thm:hc-surely-lazyD}
\end{lemma}
\begin{proof}
	See the Agda proof at the following URL:\\
	\urlLazyDProof
\end{proof}

\begin{theorem}[\lazyD{} Hypercoercion Respect \ineffCEKD]
  If $\judgetype{\emptyset}{e}{T}$ and $o : T$
  \[
  \text{\evalEqv{\effCEK{H}}{\ineffCEKD}}
  \]
\end{theorem}
\begin{proof}
  Immediately from that every Lazy D Cast ADT is correct ( 
  Theorem~\ref{thm:surely-lazyD-correct}) and that Lazy D hypercoercion is a 
  Lazy D Cast ADT (Lemma~\ref{thm:hc-surely-lazyD}). 
  %
  Alternatively, from Lemma~\ref{thm:surely-monoidic-eval},
  Lemma~\ref{thm:hc-surely-lazyD}, Proposition~\ref{thm:LazyD-HC-monoid}, and 
  Proposition~\ref{thm:LazyD-HC-identity-casts}
\end{proof}

@@

\todo[inline]{
Back out and be like remember all our contribution. Bring them back to the 
beginning. Repeat but not repeat literally.
I can assume that people have eaten meat.
}
